Modeling Bandwagoning and Hype in the MLB

### 2020-09-01 Paper_Outline.Rmd
added more to paper

fig1_dat %>%
  ggplot(aes(x = as.Date(date), y = googlehits, group = team)) + 
  geom_line(aes(color = team)) + 
  labs(title="Google Trend of LA Dodgers and SF Giants (2004-2020)", x = "Date", y = "Google Hits") + 
  scale_color_manual(values=c("dodgerblue", "darkorange")) +
  theme_minimal()+
  annotate(geom="text",x=as.Date("2012-10-01"), y=33,label="SF Giants", col = "darkorange", fontface = "bold", size = 3.5) + 
  annotate(geom="text",x=as.Date("2017-01-01"), y=52,label="LA Dodgers", col = "dodgerblue", fontface = "bold", size = 3.5) + 
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_text(angle = 0, vjust = 0.5),
        panel.border = element_rect(colour = "gray85", fill = NA))
```

@@

fig2a_dat %>%
  ggplot(aes(x = as.Date(date), y = googlehits, group = team)) + 
  geom_line(aes(color = team)) + 
  theme_minimal() +
  labs(title="Google Trend of LA Dodgers and SF Giants (2004-2020)", x = "Date", y = "Google Hits") + 
  scale_color_manual(values=c("dodgerblue", "darkorange")) +
  annotate(geom="text",x=as.Date("2012-10-01"), y=45,label="SF Giants", col = "darkorange", fontface = "bold", size = 3.5) + 
  annotate(geom="text",x=as.Date("2017-01-01"), y=80,label="LA Dodgers", col = "dodgerblue", fontface = "bold", size = 3.5) + 
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_text(angle = 0, vjust = 0.5),
        panel.border = element_rect(colour = "gray85", fill = NA))
```

@@

fig2b_dat %>%
  ggplot(aes(x = as.Date(date), y = googlehits, group = team)) + 
  geom_line(aes(color = team)) + 
  theme_minimal() +
  labs(title="Google Trend of LA Dodgers and CHI Cubs (2004-2020)", x = "Date", y = "Google Hits") + 
  scale_color_manual(values=c("firebrick2", "dodgerblue")) +
  annotate(geom="text",x=as.Date("2015-06-01"), y=50,label="CHI Cubs", col = "firebrick2", fontface = "bold", size = 3.5) + 
  annotate(geom="text",x=as.Date("2019-01-01"), y=70,label="LA Dodgers", col = "dodgerblue", fontface = "bold", size = 3.5) + 
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        axis.title.y = element_text(angle = 0, vjust = 0.5),
        axis.title.x = element_blank(),
        panel.border = element_rect(colour = "gray85", fill = NA))

@@

fig3_dat %>%
  ggplot(aes(x = googlehits)) + 
  geom_histogram(bins = 20,
                 col = "azure4",
                 fill = "azure3") +
  labs(title = "Google Trend Hits for All MLB Teams", x = "Google Hits") + 
  scale_y_continuous(expand = c(0,0), limits = c(0, 4300)) +
  theme(axis.title.y=element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "gray85", fill = NA),
        panel.background = element_rect(fill = "white")
        ) 
```

@@

As seen in Figure 3a and 3b, the skew is too extreme to address with any common data 
transformations.  However, we assert that the raw data values are not meaningful in 
and of themselves, as they correspond to an arbitrary scale set by the most publicized 
month of baseball during the time window of the data collection.  Therefore, we opted 
for a data transformation that retains the rank-ordering of the observations, but not the 
scaling.  Google Hits values were quantile normalized (reference) before model building.

@@

# giants - quantile normalized hits plot for last 5 years
fig4_dat %>%
  filter(team == "San Francisco Giants",
         date <= as.Date("2020-07-01") & date >= as.Date("2015-07-01")) %>%
  ggplot(aes(x = date, y = qn_ghits)) +
  geom_line(col = "darkorange") +
  labs(title = "Quantile-Normalized Google Trend Hits for SF Giants (Past 5 Years)") + 
  theme(plot.title = element_text(face = "bold", size = 13),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(colour = "gray85"),
        panel.grid.minor.y = element_line(colour = "gray85"),
        panel.border = element_rect(colour = "gray85", fill = NA),
        panel.background = element_rect(fill = "white"),
        plot.background = element_rect(fill = "darkorange"))

# giants - monthly average hits over ALL years with phases
fig4_dat %>%
  filter(team == "San Francisco Giants") %>% 
  group_by(month) %>%
  summarise(monthly_avg = mean(googlehits)) %>%
  ggplot(aes(x = as.factor(month), y = monthly_avg, group = 1)) +
  geom_point(col = "darkorange") +
  geom_line(col = "darkorange") + 
  geom_vline(xintercept = c(4,10), linetype = "dashed", color = "black") + 
  annotate(geom = "text", x=2, y=6.5, label = "Off-Season") +
  annotate(geom = "text", x=4, y=6.5, label = "Opening Day\n", angle = 90) + 
  annotate(geom = "text", x=7, y=6.5, label = "Season") +
  annotate(geom = "text", x=10, y=6.5, label = "Post-Season\n", angle = 90) +
  annotate(geom = "text", x=11.5, y=6.5, label = "Off-Season") +
  labs(title = "Monthly Average of Google Trend Hits for SF Giants (2004-2020)") + 
  scale_x_discrete(labels = month.abb[c(1:12)]) + 
  scale_y_discrete(expand = c(0,0), limits = factor(0:9)) +
  theme(
      # plot.title = element_text(face = "bold"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(colour = "gray90"),
        panel.grid.minor.y = element_blank(),
        panel.border = element_rect(colour = "gray75", fill = NA),
        panel.background = element_rect(fill = "white")
        # plot.background = element_rect(fill = "darkorange")
  )

@@

```{r}
# Mariners - quantile normalized hits plot for last 5 years
fig4_dat %>%
  filter(team == "Seattle Mariners",
         date <= as.Date("2020-07-01") & date >= as.Date("2015-07-01")) %>%
  ggplot(aes(x = date, y = qn_ghits)) +
  geom_line(col = "cyan4") +
  labs(title = "Quantile-Normalized Google Trend Hits for SEA Mariners (Past 5 Years)") + 
  theme(plot.title = element_text(face = "bold", size = 13),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(colour = "gray85"),
        panel.grid.minor.y = element_line(colour = "gray85"),
        panel.border = element_rect(colour = "gray85", fill = NA),
        panel.background = element_rect(fill = "white"),
        plot.background = element_rect(fill = "cyan4"))

# Mariners - monthly average hits over ALL years with phases
fig4_dat %>%
  filter(team == "Seattle Mariners") %>% 
  group_by(month) %>%
  summarise(monthly_avg = mean(googlehits)) %>%
  ggplot(aes(x = as.factor(month), y = monthly_avg, group = 1)) +
  geom_point(col = "cyan4") +
  geom_line(col = "cyan4") + 
  geom_vline(xintercept = c(4,10), linetype = "dashed", color = "black") + 
  annotate(geom = "text", x=2, y=1.75, label = "Off-Season") +
  annotate(geom = "text", x=4, y=1.75, label = "Opening Day\n", angle = 90) + 
  annotate(geom = "text", x=7, y=1.75, label = "Season") +
  annotate(geom = "text", x=10, y=1.75, label = "Post-Season\n", angle = 90) +
  annotate(geom = "text", x=11.5, y=1.75, label = "Off-Season") +
  labs(title = "Monthly Average of Google Trend Hits for SEA Mariners (2004-2020)") + 
  scale_x_discrete(labels = month.abb[c(1:12)]) + 
  scale_y_continuous(expand = c(0,0), limits = c(0, 3.5)) +
  theme(
        # plot.title = element_text(face = "bold"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(colour = "gray90"),
        panel.grid.minor.y = element_line(colour = "gray90"),
        panel.border = element_rect(colour = "gray80", fill = NA),
        panel.background = element_rect(fill = "white")
        # plot.background = element_rect(fill = "cyan4")
        )

@@

```{r}
# ALL TEAMS - quantile normalized hits plot for last 5 years
fig4_dat %>%
  filter(date <= as.Date("2020-07-01") & date >= as.Date("2015-07-01")) %>%
  group_by(date) %>%
  summarise(avg_qn_ghits = mean(qn_ghits)) %>%
  ggplot(aes(x = date, y = avg_qn_ghits)) +
  geom_line(col = "azure4") +
  labs(title = "Average Quantile-Normalized Google Trend Hits for MLB Teams (Past 5 Years)") + 
  theme(plot.title = element_text(face = "bold", size = 13),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(colour = "gray85"),
        panel.grid.minor.y = element_line(colour = "gray85"),
        panel.border = element_rect(colour = "gray85", fill = NA),
        panel.background = element_rect(fill = "white"),
        plot.background = element_rect(fill = "azure3"))

# ALL TEAMS - monthly average hits over ALL years with phases
fig4_dat %>%
  group_by(month) %>%
  summarise(monthly_avg = mean(googlehits)) %>%
  ggplot(aes(x = as.factor(month), y = monthly_avg, group = 1)) +
  geom_point(col = "azure4") +
  geom_line(col = "azure4") + 
  geom_vline(xintercept = c(4,10), linetype = "dashed", color = "black") + 
  annotate(geom = "text", x=2, y=2.5, label = "Off-Season") +
  annotate(geom = "text", x=4, y=2.5, label = "Opening Day\n", angle = 90) + 
  annotate(geom = "text", x=7, y=2.5, label = "Season") +
  annotate(geom = "text", x=10, y=2.5, label = "Post-Season\n", angle = 90) +
  annotate(geom = "text", x=11.5, y=2.5, label = "Off-Season") +
  labs(title = "Monthly Average of Google Trend Hits for MLB Teams (2004-2020)") + 
  scale_x_discrete(labels = month.abb[c(1:12)]) + 
  scale_y_continuous(expand = c(0,0), limits = c(0, 5)) +
  theme(plot.title = element_text(face = "bold"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(colour = "gray85"),
        panel.grid.minor.y = element_blank(),
        panel.border = element_rect(colour = "gray85", fill = NA),
        panel.background = element_rect(fill = "white")
        # plot.background = element_rect(fill = "azure3")
        )

@@

@@

>>>>

fig1_dat %>%
  ggplot(aes(x = as.Date(date), y = googlehits, group = team)) + 
  geom_line(aes(color = team)) + 
  labs(title="Google Trend of LA Dodgers and SF Giants (2004-2020)", x = "Date", y = "Google Hits") + 
  scale_color_manual(values=c("dodgerblue", "darkorange")) +
  theme_minimal()+
  annotate(geom="text",x=as.Date("2012-10-01"), y=33,label="SF Giants", col = "darkorange", fontface = "bold", size = 3.5) + 
  annotate(geom="text",x=as.Date("2017-01-01"), y=52,label="LA Dodgers", col = "dodgerblue", fontface = "bold", size = 3.5) + 
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(colour = "gray92"),
        panel.grid.minor.y = element_line(colour = "gray92"),
        axis.title.x = element_blank(),
        axis.title.y = element_text(angle = 0, vjust = 0.5),
        panel.border = element_rect(colour = "gray80", fill = NA))
```

@@

fig2a_dat %>%
  ggplot(aes(x = as.Date(date), y = googlehits, group = team)) + 
  geom_line(aes(color = team)) + 
  theme_minimal() +
  labs(title="Google Trend of LA Dodgers and SF Giants (2004-2020)", x = "Date", y = "Google Hits") + 
  scale_color_manual(values=c("dodgerblue", "darkorange")) +
  annotate(geom="text",x=as.Date("2012-10-01"), y=45,label="SF Giants", col = "darkorange", fontface = "bold", size = 3.5) + 
  annotate(geom="text",x=as.Date("2017-01-01"), y=80,label="LA Dodgers", col = "dodgerblue", fontface = "bold", size = 3.5) + 
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(colour = "gray92"),
        panel.grid.minor.y = element_line(colour = "gray92"),
        axis.title.x = element_blank(),
        axis.title.y = element_text(angle = 0, vjust = 0.5),
        panel.border = element_rect(colour = "gray80", fill = NA))
```

@@

fig2b_dat %>%
  ggplot(aes(x = as.Date(date), y = googlehits, group = team)) + 
  geom_line(aes(color = team)) + 
  theme_minimal() +
  labs(title="Google Trend of LA Dodgers and CHI Cubs (2004-2020)", x = "Date", y = "Google Hits") + 
  scale_color_manual(values=c("firebrick2", "dodgerblue")) +
  annotate(geom="text",x=as.Date("2015-06-01"), y=50,label="CHI Cubs", col = "firebrick2", fontface = "bold", size = 3.5) + 
  annotate(geom="text",x=as.Date("2019-01-01"), y=70,label="LA Dodgers", col = "dodgerblue", fontface = "bold", size = 3.5) + 
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(colour = "gray92"),
        panel.grid.minor.y = element_line(colour = "gray92"),
        axis.title.y = element_text(angle = 0, vjust = 0.5),
        axis.title.x = element_blank(),
        panel.border = element_rect(colour = "gray80", fill = NA))

@@

fig3_dat %>%
  ggplot(aes(x = googlehits)) + 
  geom_histogram(bins = 20,
                 col = "azure4",
                 fill = "azure3") +
  labs(title = "Google Trend Hits for All MLB Teams", x = "Google Hits") + 
  scale_y_continuous(expand = c(0,0), limits = c(0, 4300)) +
  theme(axis.title.y=element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "gray80", fill = NA),
        panel.background = element_rect(fill = "white")
        ) 
```

@@

As seen in Figures 3a and 3b, the skew is too extreme to address with any common data 
transformations.  However, we assert that the raw data values are not meaningful in 
and of themselves, as they correspond to an arbitrary scale set by the most publicized 
month of baseball during the time window of the data collection.  Therefore, we opted 
for a data transformation that retains the rank-ordering of the observations, but not the 
scaling.  Google Hits values were quantile normalized (reference) before model building.

@@

# giants - quantile normalized hits plot for last 5 years
fig4_dat %>%
  filter(team == "San Francisco Giants",
         date <= as.Date("2020-07-01") & date >= as.Date("2015-07-01")) %>%
  ggplot(aes(x = date, y = qn_ghits)) +
  geom_line(col = "darkorange") +
  labs(title = "Quantile-Normalized Google Trend Hits for SF Giants (Past 5 Years)") + 
  theme(
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(colour = "gray92"),
        panel.grid.minor.y = element_line(colour = "gray92"),
        panel.border = element_rect(colour = "gray80", fill = NA),
        panel.background = element_rect(fill = "white")
  )

# giants - monthly average hits over ALL years with phases
fig4_dat %>%
  filter(team == "San Francisco Giants") %>% 
  group_by(month) %>%
  summarise(monthly_avg = mean(googlehits)) %>%
  ggplot(aes(x = as.factor(month), y = monthly_avg, group = 1)) +
  geom_point(col = "darkorange") +
  geom_line(col = "darkorange") + 
  geom_vline(xintercept = c(4,10), linetype = "dashed", color = "black") + 
  annotate(geom = "text", x=2, y=6.5, label = "Off-Season") +
  annotate(geom = "text", x=4, y=6.5, label = "Opening Day\n", angle = 90) + 
  annotate(geom = "text", x=7, y=6.5, label = "Season") +
  annotate(geom = "text", x=10, y=6.5, label = "Post-Season\n", angle = 90) +
  annotate(geom = "text", x=11.5, y=6.5, label = "Off-Season") +
  labs(title = "Monthly Average Google Trend Hits for SF Giants (2004-2020)") + 
  scale_x_discrete(labels = month.abb[c(1:12)]) + 
  scale_y_continuous(expand = c(0,0), limits = c(0,10)) + 
  theme(
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(colour = "gray92"),
        panel.grid.minor.y = element_blank(),
        panel.border = element_rect(colour = "gray80", fill = NA),
        panel.background = element_rect(fill = "white")
  )

@@

```{r}
# Mariners - quantile normalized hits plot for last 5 years
fig4_dat %>%
  filter(team == "Seattle Mariners",
         date <= as.Date("2020-07-01") & date >= as.Date("2015-07-01")) %>%
  ggplot(aes(x = date, y = qn_ghits)) +
  geom_line(col = "cyan4") +
  labs(title = "Quantile-Normalized Google Trend Hits for SEA Mariners (Past 5 Years)") + 
  theme(
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(colour = "gray92"),
        panel.grid.minor.y = element_line(colour = "gray92"),
        panel.border = element_rect(colour = "gray80", fill = NA),
        panel.background = element_rect(fill = "white")
        )

# Mariners - monthly average hits over ALL years with phases
fig4_dat %>%
  filter(team == "Seattle Mariners") %>% 
  group_by(month) %>%
  summarise(monthly_avg = mean(googlehits)) %>%
  ggplot(aes(x = as.factor(month), y = monthly_avg, group = 1)) +
  geom_point(col = "cyan4") +
  geom_line(col = "cyan4") + 
  geom_vline(xintercept = c(4,10), linetype = "dashed", color = "black") + 
  annotate(geom = "text", x=2, y=1.75, label = "Off-Season") +
  annotate(geom = "text", x=4, y=1.75, label = "Opening Day\n", angle = 90) + 
  annotate(geom = "text", x=7, y=1.75, label = "Season") +
  annotate(geom = "text", x=10, y=1.75, label = "Post-Season\n", angle = 90) +
  annotate(geom = "text", x=11.5, y=1.75, label = "Off-Season") +
  labs(title = "Monthly Average Google Trend Hits for SEA Mariners (2004-2020)") + 
  scale_x_discrete(labels = month.abb[c(1:12)]) + 
  scale_y_continuous(expand = c(0,0), limits = c(0,4)) +
  theme(
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(colour = "gray92"),
        panel.grid.minor.y = element_blank(),
        panel.border = element_rect(colour = "gray80", fill = NA),
        panel.background = element_rect(fill = "white")
        )

@@

```{r}
# ALL TEAMS - quantile normalized hits plot for last 5 years
fig4_dat %>%
  filter(date <= as.Date("2020-07-01") & date >= as.Date("2015-07-01")) %>%
  group_by(date) %>%
  summarise(avg_qn_ghits = mean(qn_ghits)) %>%
  ggplot(aes(x = date, y = avg_qn_ghits)) +
  geom_line(col = "azure4") +
  labs(title = "Average Quantile-Normalized Google Trend Hits for MLB Teams (Past 5 Years)") + 
  theme(
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(colour = "gray92"),
        panel.grid.minor.y = element_line(colour = "gray92"),
        panel.border = element_rect(colour = "gray80", fill = NA),
        panel.background = element_rect(fill = "white")
        )

# ALL TEAMS - monthly average hits over ALL years with phases
fig4_dat %>%
  group_by(month) %>%
  summarise(monthly_avg = mean(googlehits)) %>%
  ggplot(aes(x = as.factor(month), y = monthly_avg, group = 1)) +
  geom_point(col = "azure4") +
  geom_line(col = "azure4") + 
  geom_vline(xintercept = c(4,10), linetype = "dashed", color = "black") + 
  annotate(geom = "text", x=2, y=2.5, label = "Off-Season") +
  annotate(geom = "text", x=4, y=2.5, label = "Opening Day\n", angle = 90) + 
  annotate(geom = "text", x=7, y=2.5, label = "Season") +
  annotate(geom = "text", x=10, y=2.5, label = "Post-Season\n", angle = 90) +
  annotate(geom = "text", x=11.5, y=2.5, label = "Off-Season") +
  labs(title = "Monthly Average Google Trend Hits for MLB Teams (2004-2020)") + 
  scale_x_discrete(labels = month.abb[c(1:12)]) + 
  scale_y_continuous(expand = c(0,0), limits = c(0, 5.5)) +
  theme(
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(colour = "gray92"),
        panel.grid.minor.y = element_blank(),
        panel.border = element_rect(colour = "gray80", fill = NA),
        panel.background = element_rect(fill = "white")
        )

@@

MLB teams that have done well since 2004, like the San Francisco Giants, will typically have higher quantile-normalized Google trend hits over the past 5 years than the average MLB team.  Moreover, the Giants had a higher monthly average Google hits than the average team; they even had higher spikes during the post-season because they were very successful during this time.  On the other hand, an MLB team who is less successful, such as the Seattle Mariners, will have lower quantile-normalized Google trends over the past 5 years and a lower monthly average Google hits than the average team.  The Mariners monthly average Google hits also severely drops during the post-season; a large contribution to this is that they have never made the World Series.

@@

As mentioned, the data was quantile-normaized before construction of the model began.
